Human Factors Influencing Visual Statistical Inference

### 2013-06-07 sociology_paper.tex

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis.  What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. (Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generate as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, qpproximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true then the actual data plot should look just like any of the other plots, and not be identified by an observer. %A party needs 270 electoral votes or more to win the presidential election.

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{tower.pdf} 
   \caption{ \label{fig:tower} \dc{Which one of the plots is the most from the others? Maybe this question will have people picking 4, because it is the only one with red higher.}}
\end{figure}

>>>>

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{electoral-5-13.pdf} 
   \caption{ \label{fig:elect-1} \dc{Which one of the plots is the most different from the others? Maybe this question will have people picking 4, because it is the only one where the red side is higher.}}
\end{figure}

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis. 
, indicating that we have enough evidence to reject the null hypothesis.
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' \citet{nytimes} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generate as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true then the actual data plot should look just like any of the other plots, and not be identified by an observer.
 \hh{Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.}
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Freedom's Lighthouse Averages \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   RealClearPolitics Averages\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right a histogram of 10,000 simulations from the same model. The $p$-value of observing results as extreme as in the  2012 U.S. election based on the bootstrap is 0.0533 (with Bootstrap standard error of 0.002).}
\end{figure}
